Linear stability of magnetized massive protoplanetary disks

### 2014-05-29 paper/apj/response.txt

I thank the reviewer for comments and questions. I have performed
additional calculations and extended the discussion section to address
the issues raised below. Revisions to the manuscript are made in
bold.

**********************************************************************

1) For a fixed value of B_z, massive disks would have larger values of
plasma beta, compared to the less massive disks. In actual disks
around young stars or compact objects, how large is the required value
of B_z in order for self-gravity to affect MRI modes?  
**********************************************************************
I have expanded Section 7 to include a quantitative discussion on the
limiting field strength to permit the MRI. I show that a larger plasma
beta is required in strongly self-gravitating disks. I also estimate
the upper limit to the field strength, B_z, for Q<<1.

In addition, Fig. 2 in Section 4.1.1 has been updated to include a new
calculation, \beta=25 and \Lambda_0=10. This case demonstrates the
stabilizing effect of self-gravity in the ideal limit. 
**********************************************************************

2) The amplitude of the resistivity variation from the disk surface to
the midplane (A=eta(0) / eta(infinity) ) is fixed to 100. If you take
larger values of A, such as A=10^3 or 10^4, as seen in Figure 1 of
Gressel et al. (2012), is there a qualitative change in the result?
For example, in the Q=0.2 case in Figure 5, does the magnetic energy
still penetrate into the resistive zone? In a massive disk, the value
of A is expected to be large. Could you comment on the qualitative
tendency for high A?  
**********************************************************************
Additional calculations with A=10^3 and A=10^4 are described in
Section 4.1.2. Similar results were obtained for large A with small
Lambda_0. When the midplane resistivity (or Lambda_0) is fixed and A
is increased, the magnetic energy in the resistive zone lessens. For
A=10^3 or A=10^4 and Lambda_0=0.1, both self-gravitating and
non-self-gravitating disks show little magnetic energy in the
resistive zones. However, lowering Lambda_0 with these larger values
of A yields similar results to the case in Fig. 5.

**********************************************************************
3) Section 3.1.1 Case 1: vertical field
Please mention that the odd modes are considered in Section 6, and
cite references for explaining how the boundary conditions, Equation
(29), are derived, e.g., Sano and Miyama (1999). 
**********************************************************************
Section 3.1 has been re-structured. I describe where in the results
sections are the different field setups considered. I also make the
reference to Sano & Miyama (1999).

**********************************************************************
4) Caption of Figure 11
Q=0.14 (bottom) --> Q=0.12 (bottom) 
**********************************************************************
Corrected.
**********************************************************************

### 2014-05-30 paper/linear.tex
final edits

\subsection{Domain and boundary conditions}\label{domain}
{\bf 
For a vertical field, considered in \S\ref{result1} and
\S\ref{result2}, we take $\bm{U}$ to be an even function of
$z$. Odd modes are permitted in \S\ref{result3}, where an azimuthal field may
be included. In both
setups the gravitational potential boundary condition, given by
\cite{goldreich65a}, is
%The potential boundary
%condition is 
%\begin{align}
%  \dphi^\prime(\zmax) + k_x\dphi(\zmax) = 0 \label{pot_bc1},
%\end{align}
%and is adapted from \cite{goldreich65a}. The general form of this
%potential boundary condition is used for case 2 below.
%and 
%the potential boundary condition is 
\begin{align}
  \dphi^\prime(\pm \zmax) \pm k_x\dphi(\pm\zmax) =  - \left. \frac{\Omega^2
    \rho \xi_z}{\rho_0Q}\right|_{\pm\zmax}, \label{pot_bc2}
\end{align}
where $\xi_z=\dvz/\imgi\sigma$ is the vertical Lagrangian
displacement, and $z=\pm Z_s$ is the upper and lower disk surfaces,
respectively. 
}
%In practice,  Eq. \ref{pot_bc1} and \ref{pot_bc2} are
%very similar because $\rho(\pm\zmax)\ll\rho_0$. 
%}

\subsubsection{Case 1: vertical field}
{\bf Here we impose} $d\bm{U}/dz=0$ at $z=0$. This permits higher
numerical resolution by reducing the computational domain to
$z\in~[0,Z_s]$. At the upper disk boundary $z=Z_s$ we set  
\begin{align}
  \dbx(Z_s) = \dby(Z_s) = \dvz (Z_s) = 0, 
\end{align}
so the field remains vertical. {\bf The derivation of the magnetic
  field boundary conditions may be found in \cite{sano99}.}

\subsubsection{Case 2: tilted field} 
In this more general setup the computational domain is
$z\in[-Z_s, Z_s]$ {\bf and no symmetry across the midplane is
  enforced.} At the disk surfaces we adopt the `halo' model of
\cite{gammie94}, so that
\begin{align}
  \Delta(\pm Z_s) & = 0, \\
  \dby(\pm Z_s) & = 0,\\
  \dbz(\pm Z_s) \mp \imgi \dbx(\pm Z_s) &= 0,
\end{align}
    {\bf and this case permits $\dd v_z(\pm Z_s)\neq0$.}
%      In practice, the gravitational potential boundary condition is
%      similar to case 1 because $\rho(\pm Z_s)\ll\rho_0$.}

>>>>

\subsection{Domain and boundary conditions}\label{domain}

For a vertical field, considered in \S\ref{result1} and
\S\ref{result2}, we take $\bm{U}$ to be an even function of
$z$. Odd modes are permitted in \S\ref{result3}, where an azimuthal field may
be included. In both
setups the gravitational potential boundary condition, given by
\cite{goldreich65a}, is
%The potential boundary
%condition is 
%\begin{align}
%  \dphi^\prime(\zmax) + k_x\dphi(\zmax) = 0 \label{pot_bc1},
%\end{align}
%and is adapted from \cite{goldreich65a}. The general form of this
%potential boundary condition is used for case 2 below.
%and 
%the potential boundary condition is 
\begin{align}
  \dphi^\prime(\pm \zmax) \pm k_x\dphi(\pm\zmax) =  - \left. \frac{\Omega^2
    \rho \xi_z}{\rho_0Q}\right|_{\pm\zmax}, \label{pot_bc2}
\end{align}
where $\xi_z=\dvz/\imgi\sigma$ is the vertical Lagrangian
displacement, and $z=\pm Z_s$ is the upper and lower disk surfaces,
respectively.

%In practice,  Eq. \ref{pot_bc1} and \ref{pot_bc2} are
%very similar because $\rho(\pm\zmax)\ll\rho_0$. 
%}

\subsubsection{Case 1: vertical field}
 Here we impose $d\bm{U}/dz=0$ at $z=0$. This permits higher
numerical resolution by reducing the computational domain to
$z\in~[0,Z_s]$. At the upper disk boundary $z=Z_s$ we set  
\begin{align}
  \dbx(Z_s) = \dby(Z_s) = \dvz (Z_s) = 0, 
\end{align}
so the field remains vertical. The derivation of the magnetic
  field boundary conditions may be found in \cite{sano99}.

\subsubsection{Case 2: tilted field} 
In this more general setup the computational domain is
$z\in[-Z_s, Z_s]$ and no symmetry across the midplane is
  enforced. At the disk surfaces we adopt the `halo' model of
\cite{gammie94}, so that
\begin{align}
  \Delta(\pm Z_s) & = 0, \\
  \dby(\pm Z_s) & = 0,\\
  \dbz(\pm Z_s) \mp \imgi \dbx(\pm Z_s) &= 0,
\end{align}
     and this case permits $\dd v_z(\pm Z_s)\neq0$.
%      In practice, the gravitational potential boundary condition is
%      similar to case 1 because $\rho(\pm Z_s)\ll\rho_0$.}

### 2014-05-30 paper/paper.tex
final edits

\begin{abstract}
  Magneto-rotational instability (MRI) and gravitational instability
  (GI) are the two principle routes to turbulent angular momentum
  transport in accretion disks. Protoplanetary disks may develop 
  both. This paper aims to reinvigorate interest in the study of
  magnetized  massive protoplanetary disks, starting from the basic
  issue of stability. The local linear stability of a 
  self-gravitating, uniformly magnetized, differentially rotating, 
  three-dimensional disk subject to axisymmetric perturbations is
  calculated numerically. The formulation includes 
  resistivity. It is found that the reduction in the disk thickness
  by self-gravity can decrease MRI growth rates; the MRI
  becomes global in the vertical direction, and MRI modes 
  with small radial length scales are stabilized. In massive
  disks with layered resistivity, the MRI is not well-localized
  to regions where the Elsasser number exceeds unity. For MRI modes
  with radial length scales on the order of the disk thickness, self-gravity 
  can enhance density perturbations, an effect that becomes 
  significant in the presence of a strong toroidal field. In
  gravitationally unstable disks where GI and MRI growth rates are
  comparable, the character of unstable modes can transition smoothly
  between MRI and GI. Implications for non-linear simulations are
  discussed briefly.

@@

>>>>

\begin{abstract}
  Magneto-rotational instability (MRI) and gravitational instability
  (GI) are the two principle routes to turbulent angular momentum
  transport in accretion disks. Protoplanetary disks may develop 
  both. This paper aims to reinvigorate interest in the study of
  magnetized  massive protoplanetary disks, starting from the basic
  issue of stability. The local linear stability of a 
  self-gravitating, uniformly magnetized, differentially rotating, 
  three-dimensional stratified disk subject to axisymmetric perturbations is
  calculated numerically. The formulation includes 
  resistivity. It is found that the reduction in the disk thickness
  by self-gravity can decrease MRI growth rates; the MRI
  becomes global in the vertical direction, and MRI modes 
  with small radial length scales are stabilized. The
    maximum vertical field strength that permits the MRI
    in a strongly self-gravitating polytropic disk  with polytropic index $\Gamma=1$ 
    is estimated to be $B_{z,\mathrm{max}} \simeq \csmid\Omega\sqrt{\mu_0/16\pi G} $, where $\csmid$ is the midplane sound speed and $\Omega$ is the angular velocity. 
  In massive
  disks with layered resistivity, the MRI is not well-localized
  to regions where the Elsasser number exceeds unity. For MRI modes
  with radial length scales on the order of the disk thickness, self-gravity 
  can enhance density perturbations, an effect that becomes 
  significant in the presence of a strong toroidal field, and
    which depends on the symmetry of the underlying MRI mode. In
  gravitationally unstable disks where GI and MRI growth rates are
  comparable, the character of unstable modes can transition smoothly
  between MRI and GI. Implications for non-linear simulations are
  discussed briefly.

@@

\acknowledgements
I thank K. Menou, S. Fromang and A. Youdin for helpful discussions
during the course of this project.

### 2014-05-30 paper/results_gi.tex
final edits

The GI and MRI branches only interact when their growth rates
are similar. This is seen in Fig. \ref{compare_growth3} for $Q=~0.18$
where the GI branch approaches a MRI branch at $\beta\simeq 25,\,
\gamma\simeq 0.2\Omega$. In fact, following the red curve to smaller
$\beta$ indicates GI transitions to MRI. The
`gaps' in the GI and MRI branches for $Q=0.18$ and $Q=0.12$ may be due
to the phenomenon of avoided crossing, as seen in stars
\citep[e.g.][]{aizenman77} and accretion tori{\bf /disks
\citep[e.g.][]{christo93,ogilvie98}}, where physically distinct modes approach
one another in frequency and exchange character. However, we cannot
exclude the possibility  that some modes may have been missed in a 
numerical search of eigenfrequencies.  
%maybe ogilvie?

@@

The dispersion relation $\gamma(k_x)$ deviates from the
pure GI/MRI curves with increasing $k_x$, implying stronger interaction between
magnetic and density perturbations. Comparing pure GI (dashed line)
and the gravitationally-dominated portions of $\gamma(k_x)$ shows that 
inclusion of magnetic field stabilizes high-$k_x$ pure GI. (Note also
the slight decrease in the most unstable $k_x$.) 
This stabilization is due to magnetic pressure \citep{lizano10}, {\bf 
consistent with pressure stabilizing small-wavelength GI only.}

Comparing pure MRI (solid line) and the magnetically-dominated
portions of $\gamma(k_x)$ show that self-gravity increases MRI growth
rates at large $k_x$. This effect is small but noticeable, which can
be used as a code test for non-linear simulations. Note that this
destabilization is through {\bf self-gravity in} the linear response, rather 
than through the background stratification (which is stabilizing).

>>>>

The GI and MRI branches only interact when their growth rates
are similar. This is seen in Fig. \ref{compare_growth3} for $Q=~0.18$
where the GI branch approaches a MRI branch at $\beta\simeq 25,\,
\gamma\simeq 0.2\Omega$. In fact, following the red curve to smaller
$\beta$ indicates GI transitions to MRI. The
`gaps' in the GI and MRI branches for $Q=0.18$ and $Q=0.12$ may be due
to the phenomenon of avoided crossing, as seen in stars
\citep[e.g.][]{aizenman77} and accretion tori/disks
\citep[e.g.][]{christo93,ogilvie98}, where physically distinct modes approach
one another in frequency and exchange character. However, we cannot
exclude the possibility  that some modes may have been missed in a 
numerical search of eigenfrequencies.  
%maybe ogilvie?

@@

The dispersion relation $\gamma(k_x)$ deviates from the
pure GI/MRI curves with increasing $k_x$, implying stronger interaction between
magnetic and density perturbations. Comparing pure GI (dashed line)
and the gravitationally-dominated portions of $\gamma(k_x)$ shows that 
inclusion of magnetic field stabilizes high-$k_x$ pure GI. (Note also
the slight decrease in the most unstable $k_x$.) 
This stabilization is due to magnetic pressure \citep{lizano10},  
consistent with pressure stabilizing small-wavelength GI only.

Comparing pure MRI (solid line) and the magnetically-dominated
portions of $\gamma(k_x)$ show that self-gravity increases MRI growth
rates at large $k_x$. This effect is small but noticeable, which can
be used as a code test for non-linear simulations. Note that this
destabilization by self-gravity is through the linear response, rather 
than through the background stratification (which is stabilizing).

### 2014-05-30 paper/setup.tex
first revision done
final revised version
final edits

\subsection{Resistivity profile}\label{resis_profile}
We adopt constant resistivity or a 
resistivity prescription such that $\eta(z)$ increases away from the
midplane. In the latter case, we follow \cite{fleming03} and use 
resistivity profile 
\begin{align}
  \eta(z) =
  \sqrt{2}\eta_0\left[\exp{\left(-g_+\right)}+\exp{\left(-g_-\right)}\right]^{-1/2},  
\end{align}
where
\begin{align}
  &g_\pm(z) =  \frac{\Sigma_\pm(z)-\Sigma_0}{\Sigma_*}, \\
  &\Sigma_\pm(z) = \int_{\pm z}^\infty\rho(z^\prime)dz^\prime, \label{sigma_pm}
\end{align}
and $\Sigma_0\equiv\Sigma_{\pm}(0)$, so that $g_\pm(0)=0$ and $\eta_0 
= \eta(0)$. The constant $\Sigma_*$ is chosen such that 
\begin{align}
  \cosh{\left(\frac{\Sigma_0}{\Sigma_*}\right)} =
  \left[\frac{\eta_0}{\eta(\infty)}\right]^2,
\end{align}
and we define $\eta_0/\eta(\infty)\equiv A$ as the conductivity 
boost factor from the midplane to the disk surface. We remark that
once $\rho$ and $d\rho/dz$ are obtained from
Eq. \ref{eqm_eqns1} --- \ref{eqm_eqns2}, the integration for
Eq. \ref{sigma_pm} can be performed implicitly by using Poisson's 
equation.

@@

The plasma $\beta$ measures the inverse strength of the  
magnetic field,  
\begin{align}
  \beta \equiv \frac{\csmid^2}{v_{A0}^2} =
  \frac{\csmid^2\mu_0\rho_0}{B_z^2},  
\end{align}
where $v_{A0}$ is the midplane Alfven speed. Note that we use 
the vertical field for this definition throughout this paper.

>>>>

\subsection{Resistivity profile}\label{resis_profile}
We adopt constant resistivity or a 
resistivity prescription such that $\eta(z)$ increase towards the
midplane. In the latter case, we follow \cite{fleming03} and use 
the resistivity profile 
\begin{align}
  \eta(z) =
  \sqrt{2}\eta_0\left[\exp{\left(-g_+\right)}+\exp{\left(-g_-\right)}\right]^{-1/2},  
\end{align}
where
\begin{align}
  &g_\pm(z) =  \frac{\Sigma_\pm(z)-\Sigma_0}{\Sigma_*}, \\
  &\Sigma_\pm(z) = \int_{\pm z}^\infty\rho(z^\prime)dz^\prime, \label{sigma_pm}
\end{align}
and $\Sigma_0\equiv\Sigma_{\pm}(0)$, so that $g_\pm(0)=0$ and $\eta_0 
= \eta(0)$. The constant $\Sigma_*$ is chosen such that 
\begin{align}
  \cosh{\left(\frac{\Sigma_0}{\Sigma_*}\right)} =
  \left[\frac{\eta_0}{\eta(\infty)}\right]^2,
\end{align}
and we define $\eta_0/\eta(\infty)\equiv A$ as the conductivity 
boost factor from the midplane to the disk surface. We remark that
once $\rho$ and $d\rho/dz$ are obtained from
Eq. \ref{eqm_eqns1} --- \ref{eqm_eqns2}, the integration for
Eq. \ref{sigma_pm} can be performed implicitly by using Poisson's 
equation.

@@

The plasma $\beta$ measures the inverse strength of the  
magnetic field 
\begin{align}
  \beta \equiv \frac{\csmid^2}{v_{A0}^2} =
  \frac{\csmid^2\mu_0\rho_0}{B_z^2},  
\end{align}
where $v_{A0}$ is the midplane Alfven speed. Note that we use 
the vertical field for this definition throughout this paper.

### 2014-05-30 paper/summary.tex
final edits

For a vertical field, the requirement for the MRI {\bf to operate} is that   
its vertical wavelength $\lambda \lesssim 2H$. 
{\bf 
%While $\lambda$ is independent of
%the strength of self-gravity $Q$ {\bf for fixed $\beta$}, 
The disk thickness $H=H(Q)$ decreases with increasing SG. This 
reduces MRI growth rates when 
$\beta$, and hence $\lambda$, is fixed.  
}
Thus, a sufficiently massive disk can potentially suppress the MRI. 
The MRI is also restricted to larger radial 
scales as $Q$ is lowered. This means that the MRI becomes more global
in self-gravitating disks.

{\bf
  The condition $\lambda < 2H$ 
  %requires $f(Q)/\sqrt{\beta}$ to
  %  be sufficiently small (see
  %  Eq. \ref{lambda_ideal}---\ref{lambda_resis}), so larger values of
  %  $\beta$ are required to  
  %  permit the MRI with increasing self-gravity, since $f(Q)$ increases
  %  with decreasing $Q$. 
  may be written more precisely as 
  \begin{align}\label{cond1}
    \frac{n}{\mathrm{min}(\Lambda_0, 1)}\frac{f(Q)}{\sqrt{\beta}} \lesssim 1,
  \end{align}
  where $n\sim 3$ is a numerical factor and
  $\mathrm{min}(\Lambda_0,1)$ accounts for the ideal and resistive
  limits (see Eq. \ref{lambda_ideal}---\ref{lambda_resis}). Since $f$
  increases with decreasing $Q$, Eq. \ref{cond1} implies
  the MRI requires larger values of $\beta$ with  
  increasing self-gravity. 
  For definiteness, consider the ideal polytropic disk. Then Eq. \ref{cond1} is
  \begin{align}\label{cond2}
    \beta^{-1/2} \lesssim
    \frac{\sqrt{15}}{4\pi}\sqrt{Q}\arccos{\left(\frac{Q}{1+Q}\right)}. 
  \end{align}
  For a non-self-gravitating disk, $Q\to\infty$ and Eq. \ref{cond2} is
  $\beta \gtrsim 16\pi^2/30\simeq 5$. For $Q\ll 1$, the condition is
  $\beta \gtrsim 64/15Q$, giving $\beta\gtrsim 20$ for
  $Q=0.2$. We confirm this numerically, finding the MRI growth rate
  $\gamma  \lesssim 0.1$ when $\beta \lesssim 3.3$ for $Q=20$ and 
  $\beta \lesssim 17$ for $Q=0.2$.  
%  Inserting $Q\sim 0.2$ for an ideal disk marginally stable to
%  self-gravity, we find $v_{A0}\lesssim 0.2\csmid$ for the MRI.

@@

}

>>>>

For a vertical field, the requirement for the MRI to operate is that   
its vertical wavelength $\lambda \lesssim 2H$.  
%While $\lambda$ is independent of
%the strength of self-gravity $Q$ {\bf for fixed $\beta$}, 
The disk thickness $H=H(Q)$ decreases with increasing SG. This 
reduces MRI growth rates when 
$\beta$, and hence $\lambda$, is fixed.  
Thus, a sufficiently massive disk can potentially suppress the MRI. 
The MRI is also restricted to larger radial 
scales as $Q$ is lowered. This means that the MRI becomes more global
in self-gravitating disks.

  The condition $\lambda < 2H$ 
  %requires $f(Q)/\sqrt{\beta}$ to
  %  be sufficiently small (see
  %  Eq. \ref{lambda_ideal}---\ref{lambda_resis}), so larger values of
  %  $\beta$ are required to  
  %  permit the MRI with increasing self-gravity, since $f(Q)$ increases
  %  with decreasing $Q$. 
  may be written more precisely as 
  \begin{align}\label{cond1}
    \frac{n}{\mathrm{min}(\Lambda_0, 1)}\frac{f(Q)}{\sqrt{\beta}} \lesssim 1,
  \end{align}
  where $n\sim 3$ is a numerical factor and
  $\mathrm{min}(\Lambda_0,1)$ accounts for the ideal and resistive
  limits (see Eq. \ref{lambda_ideal}---\ref{lambda_resis}). Since $f$
  increases with decreasing $Q$, Eq. \ref{cond1} implies
  the MRI requires larger values of $\beta$ with  
  increasing self-gravity. 
  For definiteness, consider the ideal polytropic disk. Then Eq. \ref{cond1} is
  \begin{align}\label{cond2}
    \beta^{-1/2} \lesssim
    \frac{\sqrt{15}}{4\pi}\sqrt{Q}\arccos{\left(\frac{Q}{1+Q}\right)}. 
  \end{align}
  For a non-self-gravitating disk, $Q\to\infty$ and Eq. \ref{cond2} is
  $\beta \gtrsim 16\pi^2/30\simeq 5$. For $Q\ll 1$, the condition is
  $\beta \gtrsim 64/15Q$, giving $\beta\gtrsim 20$ for
  $Q=0.2$. We confirm this numerically, finding the MRI growth rate
  $\gamma  \lesssim 0.1$ when $\beta \lesssim 3.3$ for $Q=20$ and 
  $\beta \lesssim 17$ for $Q=0.2$.  
%  Inserting $Q\sim 0.2$ for an ideal disk marginally stable to
%  self-gravity, we find $v_{A0}\lesssim 0.2\csmid$ for the MRI.

@@
